HO unification from object language to meta language

### 2024-03-26 paper/main.tex
update tex

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

@@

$\eta$-reduction is a property by which a term $t$ is of the form $\lambda x. T
x$ and $x$ is not free in $T$. In this case we case say that $T$ is equal to its
$\eta$-expanded version represented by the term $t$. Both the meta and the
object language accept terms unification up to $\eta$ equivalence, but as
sketched in the small code snippet in \cref{sec:problem-statement}, \Ue is not
able to capture this equivalence property when using terms in \Fo. The main
reason is that the unification of two terms in \Fo{} appears to be syntactical
on the constructors of these terms. As a very concise example, \Ue can't unify
the \Fo{} terms \elpiIn{lam x\ app[con "f", x]} and \elpiIn{con "f"} since the
rigid constructor \elpiIn{lam} is different from the rigid constructor
\elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
order to address this problem.

@@

\paragraph{Detection of \maybeeta} The main modification of the compiler to solve this
unification issue consists in identifying all the subterms of a term $t$ that
are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
$\eta$-expansion under a certain substitution $\sigma$, the code verifying this
property is given by the following code:

@@

\noindent
We call the two terms carried by the link respectively left and right hand side.

\begin{invariant}[\linketa left hand side]
  The left hand side of a \linketa is a variable.
  \label{invariant:link-eta-left}
\end{invariant}

\begin{invariant}[\linketa right hand side]
  The right hand side of a \linketa having the shape $\lambda x.F \appsep x$
  where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{invariant:link-eta-right}
\end{invariant}

@@

We want maintain invariants~\ref{invariant:link-eta-left} and \ref{invariant:link-eta-right}

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}\xspace}

@@

$\eta$-reduction is a property by which a term $t$ is of the form $\lambda x. T
x$ and $x$ is not free in $T$. In this case we case say that $T$ is equal to its
$\eta$-expanded version represented by the term $t$. Both the meta and the
object language accept terms unification up to $\eta$ equivalence, but as
sketched in the small code snippet in \cref{sec:problem-statement}, \Ue is not
able to capture this equivalence property when using terms in \Fo. The main
reason is that the unification of two terms in \Fo{} appears to be syntactical
on the constructors of these terms. As a very concise example, \Ue can't unify
the \Fo{} terms \elpiIn{lam x\ app[con "f", x]} and \elpiIn{con "f"} since the
rigid constructor \elpiIn{lam} is different from the rigid constructor
\elpiIn{con}. In conclusion, we need to adapt our \elpiIn{comp} relation in
order to address this problem. In the sections below, we explain how we detect
and compile \maybeeta terms with the corresponding \linketa and how these links
are used to make unification progress. The same explanation structure will be 
used for \maybebeta terms.

@@

\paragraph{Detection of \maybeeta} The main modification of the compiler to solve this
unification issue consists in identifying all the subterms of a term $t$ that
are \maybeeta. In particular, a term $t$ is a \maybeeta, if it can reduce to a
$\eta$-expansion under a certain substitution $\sigma$. The code verifying this
property is given below:

@@

\newcommand{\lhs}{\ensuremath{\mathrm{lhs}}\xspace}
\newcommand{\rhs}{\ensuremath{\mathrm{rhs}}\xspace}

\noindent
We call the two terms carried by the link respectively left and right hand side,
called respectively \lhs and \rhs.

\begin{invariant}[\linketa \lhs]
  The \lhs of a \linketa is a variable.
  \label{invariant:link-eta-left}
\end{invariant}

TODO: motivate this?

\begin{invariant}[\linketa \rhs]
  The \rhs of a \linketa having the shape $\lambda x.F \appsep x$
  where $F \appsep x$ is a term not starting with the \elpiIn{lam} constructor.
  \label{invariant:link-eta-right}
\end{invariant}

TODO: motivate this?

@@

\linketa are meant to suspend the unification of two terms, that is, if we have
a \maybeeta term $t$ which should be unified with a term $t'$ we don't want to
unify them with \Ue. As said before this would break unification but it can also
introduce \maybeeta in the substitution of \Ho.

In order to activate a \linketa, we need to implement new rules for the
\elpiIn{progress1} predicate. There are two cases making a \linketa to progress,
1) \lhs is instantiated to a rigid term, in this case \lhs is unified with \rhs.
TODO: the right eta hand side is eta-expanbded if it is an app/con; 2) \rhs can
be $\eta$-reduced to a term with rigid head, in this case \lhs and \rhs are
again unified. If one of these two condition is satisfied, the link has
fulfilled is task and can be removed from the list of suspended links; if none
the condition succeeds, the link is kept for a further iteration of
\elpiIn{progress}.

TODO: dire che scendiamo sotto i vari abs che formano il contesto?

We can note that the insertion of these rules for \elpiIn{progress1} do not
prevent the termination of \elpiIn{progress}, since, if the link is woken up,
the unification of the terms and the suppression of the link form \linkStore
are terminating operation. Note that if the link remain suspend
\elpiIn{progress} continues to terminate, since, it would mean that the
condition of the \elpiIn{if}-branchement succeds and the \elpiIn{progress}
terminates its execution.

A second way to progress \linketa is to check if \linkStore contains two
\linketa $l_1$ and $l_2$ with same \lhs. In this case we are allowed to unify

TODO: dire che termina

\begin{lemma}
  We never add eta-expansions in the substitution
\end{lemma}

\begin{proof}[Proof sketch]
\end{proof}

TODO: we can however have $\lambda x. F_x$ if we know that $F$ does not reduce
to $T x$ where x is not free in $T$.
